---
title: "covid_microbiome_analyses"
author: "Fran Querdasi"
date started: "2022-12-28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, #set cache to be saved so that knitting is faster once it has been done once
                      dev="svg", #set graphs to be saved as. svg (vector-based format rather than raster based formats like.jpg)
                      dpi=1000) #set resolution to be 1000 dpi 
#all graphs will be saved in "files" folder in same directory as script
```

# Setup
## Load libraries
```{r}
suppressPackageStartupMessages({
  library(vegan)
  library(tidyverse)
  library(stats)
  library(phyloseq)
  library(Polychrome)
  library(mefa)
  library(qiime2R)
  library(microbiome)
  library(fantaxtic)
  library(car)
  library(ggpubr)
  library(readxl)
  library(psych)
  library(sjPlot)
  library(reshape2)
  library(mosaic)
  library(summarytools) #for chi-square test and tables
  library(Maaslin2)
  library(RColorBrewer)
  library(patchwork)
  library(microViz) #for editing variables in phyloseq object
})
```

## Load files
```{r}
#for phyloseq
otu_table_path <- "../Full_Sample/shell_asv_table.qza"
species_otu_path <- "../Full_Sample/species_level/species_table.qza"
tree_path <- "../Full_Sample/tree.qza"
taxonomy_path <- "../Full_Sample/shell_taxonomy.qza"
metadata_path <- "covid_metadata_vars_test.tsv"

#for maaslin2
taxonomy_tsv <- "../Full_Sample/shell_taxonomy.tsv"
genus_tax_file <- "../Full_Sample/genus_level/genus-table.csv"
species_tax_file <- "../Full_Sample/species_level/species-table.csv"
family_tax_file <- "../Full_Sample/family_level/family-table.csv"
```

# Data wrangling
## Create phyloseq object
```{r, results='hide'}
# add pandemic string variable to metadata
meta <- read_tsv(metadata_path) %>% 
  dplyr::mutate(pandemic_str = ifelse(pandemic==0, "pre-pandemic", "pandemic"))

# write the metadata file back (name with test first)
write_tsv(meta, "covid_metadata_vars_test.tsv")

#original ASV table 
covid_phy_obj <- qza_to_phyloseq(
  features=otu_table_path,
  tree=tree_path,
  taxonomy=taxonomy_path,
  metadata=metadata_path
)

#view phyloseq object to check how it looks
sample_data(covid_phy_obj)
taxa_names(covid_phy_obj)
sample_names(covid_phy_obj)

covid_phy_obj #correct number of samples, metadata vars, etc. 
sample_variables(covid_phy_obj)
```

# Initial data visualization: taxa abundance (at species level)

```{r, warning=FALSE}
# get top 25 most abundant taxa
physeq_ttaxa <- get_top_taxa(physeq_obj = covid_phy_obj,
                             n = 25,
                             relative = TRUE,
                             discard_other = TRUE)

#group by pandemic status
physeq_ttaxa_pandemic <- merge_samples(physeq_ttaxa, "pandemic")

#create 'genus species' variable
tmp_pandemic <- name_taxa(physeq_ttaxa_pandemic, species = TRUE)

#barplot based on postnatal adversity group (NAs are removed), colors are strains, labels are species
fantaxtic_bar(tmp_pandemic, color_by = "Species", label_by = "Species") +
  xlab("Pandemic") +
  ylab("Relative Abundance") #+
  #scale_fill_manual(values=as.vector(glasbey(32)))
```

# Beta diversity analyses

```{r}
physeq_covid_analyses <- subset_samples(covid_phy_obj, pandemic != "NA" & sex != "NA" & avg_fats != "NA" & still_breastfeed != "NA" & method_delivery != "NA") #create a dataset that is complete on sex; necessary for distance matrix calculations to work (N=39 total)

nsamples(physeq_covid_analyses) #N=39

#get distance matrices (weighted unifrac, unweighted unifrac)
set.seed(5049) #need to run this right before all Unifrac distance matrices calculations 
wunifrac <- phyloseq::distance(physeq_covid_analyses, method = "wunifrac", type = "samples")
set.seed(5049) 
unifrac <- phyloseq::distance(physeq_covid_analyses, method = "unifrac", type = "samples")

#get dataframe of metadata
metadata_physeq_analyses <- data.frame(sample_data(physeq_covid_analyses)) #get data frame of metadata

# determine the number of samples in each group
metadata_physeq_analyses %>% group_by(pandemic) %>% summarize(n()) #26 pre-pandemic, 13 pandemic

# test betadispersion assumption
disp <- betadisper(wunifrac, metadata_physeq_analyses$pandemic)
permutest(disp, permutations=how(nperm=9999)) #homogeneity of dispersion is violated *****
disp_unifrac <- betadisper(unifrac, metadata_physeq_analyses$pandemic)
permutest(disp_unifrac, permutations=how(nperm=9999)) #no violation

#perform PERMANOVAs with each distance matrix (no covariates)
set.seed(5049) #need to run this before Unifrac distance matrices calculations 
adonis2(wunifrac~pandemic, data=metadata_physeq_analyses, permutations=9999)#ns
set.seed(5049)
adonis2(unifrac~pandemic, data=metadata_physeq_analyses, permutations=9999) #significant 

# add covariates
set.seed(5049)
adonis2(wunifrac~pandemic + sex + avg_fats + still_breastfeed + method_delivery, data=metadata_physeq_analyses, permutations=9999) 
set.seed(5049)
adonis2(unifrac~pandemic + sex + avg_fats + still_breastfeed + method_delivery, data=metadata_physeq_analyses, permutations=9999) #significant
```
## Visualize beta divesrity results (Figure 3)
This paper uses NMDS to visualize: https://www.researchgate.net/publication/308645859_Environment_and_host_species_shape_the_skin_microbiome_of_captive_neotropical_bats#pfb
This tutorial describes how to create ordination plots: https://joey711.github.io/phyloseq/plot_ordination-examples.html
```{r}
#change the levels of pandemic string so that pre pandemic is first before pandemic
physeq_covid_analyses_test <- physeq_covid_analyses %>% 
  ps_mutate(pandemic_str_graph = factor(pandemic_str, levels=c("pre-pandemic", "pandemic")))

#sample_data(physeq_covid_analyses_test)

# visualize weighted unifrac
ord <- ordinate(physeq_covid_analyses_test, "NMDS", "wunifrac")
p1 = plot_ordination(physeq_covid_analyses_test, ord, type="samples", color="pandemic_str_graph") +
  stat_ellipse() + 
  labs(color= "Pandemic Group") + 
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  # theme( 
  #   panel.grid.major.x = element_blank(),
  #   panel.grid.minor.x = element_blank(),
  #   panel.grid.major.y = element_blank(),
  #   panel.grid.minor.y = element_blank(),
  # ) +
  theme(legend.position = "none") +
  ggtitle('A) Weighted Unifrac')

#pre-pandemic group appears more dispersed than pandemic group, most likely because there are just very few pandemic samples

ggsave("figures/wunifrac.svg", height=4, width=6, dpi=1000)

# visualize unweighted unifrac
ord2 <- ordinate(physeq_covid_analyses_test, "NMDS", "unifrac")
p2 = 
  plot_ordination(physeq_covid_analyses_test, ord2, type="samples", color="pandemic_str_graph") +
  stat_ellipse() +
  labs(color= "Pandemic Group") + 
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  # theme( 
  #   panel.grid.major.x = element_blank(),
  #   panel.grid.minor.x = element_blank(),
  #   panel.grid.major.y = element_blank(),
  #   panel.grid.minor.y = element_blank(),
  # ) +
  ggtitle('B) Unweighted Unifrac')

#no difference in homogeneity of dispersion bw groups; significant clustering by pandemic

ggsave("figures/unifrac.svg", height=4, width=6, dpi=1000)

# combine figures into one
p1 + p2

ggsave("figures/beta_diversiy_ellipses.svg", height=4, width=6, dpi=1000)
```


# Differential abundance analyses
## Format data for differential abundance analyses
```{r}
#create genus-level taxonomy file
genus_taxonomy <- as.data.frame(read.csv(genus_tax_file))
rownames(genus_taxonomy) <- genus_taxonomy[,1]
genus_taxonomy <- genus_taxonomy[,2:ncol(genus_taxonomy)]

#transpose so taxa are on rows, samples on columns
genus_taxonomy_t <- as.data.frame(t(genus_taxonomy))
genus_taxonomy_t <- genus_taxonomy_t %>% 
  rename('#SampleID' = '#OTU ID') 
write_tsv(genus_taxonomy_t, "genus_tax_features_covid.tsv")

#create species-level taxonomy file
species_tax <- as.data.frame(read.csv(species_tax_file))
rownames(species_tax) <- species_tax[,1]
species_tax <- species_tax[,2:ncol(species_tax)]

#transpose so taxa are on rows, samples on columns
species_tax_t <- as.data.frame(t(species_tax))
species_tax_t <- species_tax_t %>% 
  rename('SampleID' = '#OTU ID') 
write_tsv(species_tax_t, "species_tax_features_covid.tsv")

#create string pandemic variable in metadata file to ensure that it is read as a categorical var
covid_metadata <- read_tsv("covid_metadata_vars.tsv") %>% 
  mutate(pandemic_str = if_else(pandemic==0, "pre", "during")) %>% 
  rename('#SampleID' = SampleID)

write_tsv(covid_metadata, "covid_metadata_vars_edited.tsv")

#how many participants have average fat data?
covid_metadata %>% dplyr::count(!is.na(avg_fats)) #41/55 have avg fat data

#create family-level taxonomy file
family_tax <- as.data.frame(read.csv(family_tax_file))
rownames(family_tax) <- family_tax[,1]
family_tax <- family_tax[,2:ncol(family_tax)]

#transpose
family_tax_t <- as.data.frame(t(family_tax))
family_tax_t <- family_tax_t %>% 
  rename('#SampleID' = '#OTU ID') 
write_tsv(family_tax_t, "family_tax_features_covid.tsv")

```
# Differential abundance analyses comparing pre-pandemic and pandemic groups
## Exploratory Analyses: genus level data
Because we have a small sample, to reduce number of comparisons we filtered via maaslin2 to only look at taxa with non-zero abundance in at least 50% of samples (27/54 complete cases) 
```{r, results='hide', warning=FALSE}
# make dataset complete on all predictors
covid_metadata <- read_tsv("covid_metadata_vars_edited.tsv")
covid_metadata_complete <- covid_metadata %>% 
  dplyr::filter(!is.na(pandemic) & !is.na(sex) & !is.na(avg_fats) & !is.na(method_delivery) & !is.na(still_breastfeed))

# get a dataset of sampleIDs
sampleids_complete <- covid_metadata_complete %>% dplyr::select('#SampleID')

genus_tax_features_covid <- read_tsv("genus_tax_features_covid.tsv")
genus_tax_features_covid_complete <- sampleids_complete %>% left_join(genus_tax_features_covid, by = "#SampleID")

#without controlling for average fats (or any covariates)
pandemic_genus <- Maaslin2(
  input_data = "genus_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_genus_filtered",
  min_abundance = 0,
  min_prevalence = 0.5, #filter to only look at taxa with 50% non-zero values (non-zero in 17/34)
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str"),
  standardize = TRUE
)

#controlling for average fat intake
pandemic_genus_covs <- Maaslin2(
  input_data = "genus_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_genus_filtered_controlfats",
  min_abundance = 0,
  min_prevalence = 0.5, #filter to only look at taxa with 50% non-zero values (non-zero in 17/34)
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str", "avg_fats"),
  standardize = TRUE
)

# R&R from scientific reports: adding child sex, birthmode, and still breastfeeding to covariates
pandemic_genus_covs <- Maaslin2(
  input_data = "genus_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_genus_filtered_covariates_r1",
  min_abundance = 0,
  min_prevalence = 0.5, #filter to only look at taxa with 50% non-zero values (non-zero in 17/34)
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str", "avg_fats", "sex", "still_breastfeed", "method_delivery"),
  standardize = TRUE
)
```

### Graph significant genera results (Figure 5)
```{r}
#create smaller dataset with pandemic string variable relevant taxa
covid_metadata_genus <- read_tsv("covid_metadata_vars_edited.tsv") %>% 
  left_join(read_tsv("genus_tax_features_covid.tsv"), by = '#SampleID') 

covid_metadata_genus <- covid_metadata_genus %>% 
  dplyr::select(
    '#SampleID',
    ID,
    dist322,
    pandemic_str,
    avg_fats,
    still_breastfeed,
    method_delivery,
    sex,
    'Bacteria;Proteobacteria;Gammaproteobacteria;Pasteurellales;Pasteurellaceae;Haemophilus',
    contains("Shigella"),
    contains("torques")
  )

#rename abundance variable for ease of plotting
covid_metadata_genus <- covid_metadata_genus %>% 
  dplyr::rename(
    g__Haemophilus = 'Bacteria;Proteobacteria;Gammaproteobacteria;Pasteurellales;Pasteurellaceae;Haemophilus',
    g__Shigella = "Bacteria;Proteobacteria;Gammaproteobacteria;Enterobacterales;Enterobacteriaceae;Escherichia-Shigella",
    g__Ruminococcus_torques = "Bacteria;Firmicutes;Clostridia;Lachnospirales;Lachnospiraceae;[Ruminococcus] torques group"   
  )

#change the levels of pandemic string so that pre pandemic is first before during
covid_metadata_genus$pandemic_str <- factor(covid_metadata_genus$pandemic_str, levels=c("pre", "during"))

# create dataset that is complete on covariates
covid_metadata_genus_complete <- covid_metadata_genus %>% dplyr::filter(!is.na(avg_fats) & !is.na(still_breastfeed) & !is.na(method_delivery) & !is.na(sex) & !is.na(pandemic_str))

#GRAPH -- Haemophilus and pandemic 
#haemophilus <- 
  ggplot(covid_metadata_genus %>% filter(!is.na(pandemic_str)), aes(y = g__Haemophilus, x = pandemic_str, color=pandemic_str)) +
geom_point(position=position_jitter(w=0.3, h=0), na.rm = TRUE) +
  geom_boxplot(alpha=0.3, outlier.shape=NA, na.rm=TRUE, data=covid_metadata_genus %>% filter(!is.na(pandemic_str))) +
  xlab("Pandemic Group") + ylab("Haemophilus Count (per sample)") + 
  scale_x_discrete("Pandemic Group", labels = c("pre" = "Pre-Pandemic", "during" = "Pandemic")) +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme( 
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
  ) +
  theme(legend.position = "none") 
  #+ ggtitle('A')

ggsave("figures/Haemophilus_pandemic.svg", height=4, width=6, dpi=1000)

# haemophilus in only complete subsample

 ggplot(covid_metadata_genus_complete %>% filter(!is.na(pandemic_str)), aes(y = g__Haemophilus, x = pandemic_str, color=pandemic_str)) +
geom_point(position=position_jitter(w=0.3, h=0), na.rm = TRUE) +
  geom_boxplot(alpha=0.3, outlier.shape=NA, na.rm=TRUE, data=covid_metadata_genus %>% filter(!is.na(pandemic_str))) +
  xlab("Pandemic Group") + ylab("Haemophilus Count (per sample)") + 
  scale_x_discrete("Pandemic Group", labels = c("pre" = "Pre-Pandemic", "during" = "Pandemic")) +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme( 
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
  ) +
  theme(legend.position = "none") 
  #+ ggtitle('A')

ggsave("figures/Haemophilus_pandemic_complete.svg", height=4, width=6, dpi=1000)
```

## Exploratory Analyses: species level data
```{r, results='hide', warning=FALSE}
#without controlling for fat intake
pandemic_species <- Maaslin2(
  input_data = "species_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_species_filtered",
  min_abundance = 0,
  min_prevalence = 0.5, #filter to only look at taxa with 50% non-zero values (non-zero in 17/34)
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str"),
  standardize = TRUE
)

#controlling for fat intake
pandemic_species_covs <- Maaslin2(
  input_data = "species_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_species_filtered_controlfats",
  min_abundance = 0,
  min_prevalence = 0.5, #filter to only look at taxa with 50% non-zero values (non-zero in 17/34)
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str", "avg_fats"),
  standardize = TRUE
)

# R&R from scientific reports: adding child sex, birth mode, and still breastfeeding to covariates
pandemic_species_covs <- Maaslin2(
  input_data = "species_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_species_filtered_covariates_r1",
  min_abundance = 0,
  min_prevalence = 0.5, #filter to only look at taxa with 50% non-zero values (non-zero in 17/34)
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str", "avg_fats", "sex", "still_breastfeed", "method_delivery"),
  standardize = TRUE
)

# still no significant differentially abundant species
```

## Targeted Analyses, family level
Families are: Bifidobacteriaceae, Lactobacillaceae, Pasteurellaceae, Lachnospiraceae, Ruminococcaceae, Tannerellaceae, Bacteroidaceae, and Streptococcaceae. These are all families that show overlap between studies of stress, hygiene, daycare attendance, household composition, and the one COVID/microbiome study published (in adults)
### Run full maasiln2 with all comparisons
```{r, results='hide', warning=FALSE}
#run full maasiln2 with all comparisons (without controlling for average fats)
pandemic_family <- Maaslin2(
  input_data = "family_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_family_targeted",
  min_abundance = 0,
  min_prevalence = 0.1, #lower filtering threshold to capture all of the families
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str"),
  standardize = TRUE
)

#run maaslin2 controlling for average fats
pandemic_family_covs <- Maaslin2(
  input_data = "family_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_family_targeted_controlfats",
  min_abundance = 0,
  min_prevalence = 0.1, #lower filtering threshold to capture all of the families
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str", "avg_fats"),
  standardize = TRUE
)

# R&R from scientific reports: adding child sex, method of delivery and still breastfeeding to covariates
pandemic_family_r1covs <- Maaslin2(
  input_data = "family_tax_features_covid.tsv",
  input_metadata = "covid_metadata_vars_edited.tsv",
  output = "da_results/pandemic_family_filtered_covariates_r1",
  min_abundance = 0,
  min_prevalence = 0.5, #filter to only look at taxa with 50% non-zero values (non-zero in 17/34)
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("pandemic_str", "avg_fats", "sex", "still_breastfeed", "method_delivery"),
  standardize = TRUE
)
```

### Calculate correct q-value on subset of families of interest
```{r}
#WITHOUT COVARIATES
#select families of interest, subset them, and recalculate q-value
#grab results of interest
da_results_families <- pandemic_family$results %>% 
filter(grepl("Bifidobacteriaceae|Lactobacillaceae|Pasteurellaceae|Lachnospiraceae|Ruminococcaceae|Tannerellaceae|Bacteroidaceae|Streptococcaceae", feature))

#generate correct q-value
da_results_families$qval<-p.adjust(da_results_families$pval, method = 'BH')

#write out the results of interest
write_csv(da_results_families, "da_results/pandemic_family_targeted/targeted_family_results.csv")

#CONTROLLING FOR AVERAGE FATS
da_results_families_controlfats <- pandemic_family_covs$results %>% 
  filter(grepl("Bifidobacteriaceae|Lactobacillaceae|Pasteurellaceae|Lachnospiraceae|Ruminococcaceae|Tannerellaceae|Bacteroidaceae|Streptococcaceae", feature))

#generate correct q-value
da_results_families_controlfats$qval<-p.adjust(da_results_families_controlfats$pval, method = 'BH')

#write out the results of interest
write_csv(da_results_families_controlfats, "da_results/pandemic_family_targeted/targeted_family_results_controlfats.csv")

# CONTROLLING FOR COVARIATES IN R1 FOR SCIENTIFIC REPORTS
da_results_families_r1covs <- pandemic_family_r1covs$results %>% 
  filter(grepl("Bifidobacteriaceae|Lactobacillaceae|Pasteurellaceae|Lachnospiraceae|Ruminococcaceae|Tannerellaceae|Bacteroidaceae|Streptococcaceae", feature)) %>% filter(metadata=="pandemic_str")

#generate correct q-value
da_results_families_r1covs$qval<-p.adjust(da_results_families_r1covs$pval, method = 'BH')

#write out the results of interest
write_csv(da_results_families_r1covs, "da_results/pandemic_family_filtered_covariates_r1/targeted_family_results_r1covs.csv")
```

### Graph Pasteurellaceae results (Figure 4)
```{r}
#create smaller dataset with pandemic string variable and Blautia, unknown species abundance
covid_metadata_family <- read_tsv("covid_metadata_vars_edited.tsv") %>% 
  left_join(read_tsv("family_tax_features_covid.tsv"), by = '#SampleID') 

covid_metadata_family <- covid_metadata_family %>% 
  mutate(
    taxa_count_per_sample = rowSums(dplyr::select(., contains("Bacteria")), na.rm=TRUE)
  )

#rename abundance variable for ease of plotting
covid_metadata_family <- covid_metadata_family %>% 
  mutate(
    f__Lactobacillaceae_perc = `Bacteria;Firmicutes;Bacilli;Lactobacillales;Lactobacillaceae`/taxa_count_per_sample,
    f__Pasteurellaceae_perc = `Bacteria;Proteobacteria;Gammaproteobacteria;Pasteurellales;Pasteurellaceae`/taxa_count_per_sample,
  ) %>% 
  dplyr::rename(
    f__Lactobacillaceae = 'Bacteria;Firmicutes;Bacilli;Lactobacillales;Lactobacillaceae',
    f__Pasteurellaceae = 'Bacteria;Proteobacteria;Gammaproteobacteria;Pasteurellales;Pasteurellaceae'
  )

#change the levels of pandemic string so that pre pandemic is first before during
covid_metadata_family$pandemic_str <- factor(covid_metadata_family$pandemic_str, levels=c("pre", "during"))

# create dataset that is complete on covariates
covid_metadata_family_complete <- covid_metadata_family %>% dplyr::filter(!is.na(avg_fats) & !is.na(still_breastfeed) & !is.na(method_delivery) & !is.na(sex) & !is.na(pandemic))

#Graph  
  ggplot(covid_metadata_family %>% filter(!is.na(pandemic_str)), aes(y = f__Pasteurellaceae, x = pandemic_str, color=pandemic_str)) +
geom_point(position=position_jitter(w=0.3, h=0), na.rm = TRUE) +
  geom_boxplot(alpha=0.3, outlier.shape=NA, na.rm=TRUE, data=covid_metadata_family %>% filter(!is.na(pandemic_str))) +
  #stat_summary(fun=mean, colour='#7570B3', geom='point', shape='-', size=15, na.rm=T) +
  xlab("Pandemic Group") + ylab("Pasteurellaceae Count (per sample)") + 
  scale_x_discrete("Pandemic Group", labels = c("pre" = "Pre-Pandemic", "during" = "Pandemic")) +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme( 
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
  ) +
  theme(legend.position = "none") 
  #+ggtitle('B')

ggsave("figures/Pasteurellaceae_prepost.svg", height=4, width=6, dpi=1000)

# pasteur with complete cases only
ggplot(covid_metadata_family_complete %>% filter(!is.na(pandemic_str)), aes(y = f__Pasteurellaceae, x = pandemic_str, color=pandemic_str)) +
geom_point(position=position_jitter(w=0.3, h=0), na.rm = TRUE) +
  geom_boxplot(alpha=0.3, outlier.shape=NA, na.rm=TRUE, data=covid_metadata_family %>% filter(!is.na(pandemic_str))) +
  #stat_summary(fun=mean, colour='#7570B3', geom='point', shape='-', size=15, na.rm=T) +
  xlab("Pandemic Group") + ylab("Pasteurellaceae Count (per sample)") + 
  scale_x_discrete("Pandemic Group", labels = c("pre" = "Pre-Pandemic", "during" = "Pandemic")) +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme( 
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
  ) +
  theme(legend.position = "none") 
  #+ggtitle('B')

ggsave("figures/pasteur_complete.svg", height=4, width=6, dpi=1000)
```

# Differential Abundance Analyses by days since pandemic onset
## Species-level 
```{r, results='hide', warning=FALSE}
#create subset of samples collected after pandemic onset
postpand <- read_tsv("covid_metadata_vars_edited.tsv") %>% dplyr::filter(dist322>0) #N=20

#write that file
write_tsv(postpand, "covid_metadata_postpand_sample.tsv")

#run full maasiln2 with all comparisons
dist322_species <- Maaslin2(
  input_data = "species_tax_features_covid.tsv",
  input_metadata = "covid_metadata_postpand_sample.tsv",
  output = "da_results/dist322_species",
  min_abundance = 0,
  min_prevalence = 0.5, 
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("dist322"),
  standardize = TRUE
)

#is average fat intake associated with days since pandemic onset?
cor.test(postpand$dist322, postpand$avg_fats) #negative association (r=-0.5), marginal (p=0.08)

postpand %>% dplyr::count(!is.na(avg_fats)) #13/20 have fat data though

ggplot(postpand, aes(x=dist322, y=avg_fats)) + 
  geom_point() +
  geom_smooth(method="lm")
#mostly driven by outlier with high fat early in pandemic and low fat later in pandemic
```

## Family-level
```{r, results='hide', warning=FALSE}
#run full maasiln2 with all comparisons
dist322_family <- Maaslin2(
  input_data = "family_tax_features_covid.tsv",
  input_metadata = "covid_metadata_postpand_sample.tsv",
  output = "da_results/dist322_family",
  min_abundance = 0,
  min_prevalence = 0.5, 
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("dist322"),
  standardize = TRUE
)

#WITHOUT COVARIATES
#select families of interest, subset them, and recalculate q-value
#grab results of interest
da_dist322_families <- dist322_family$results %>% 
filter(grepl("Bifidobacteriaceae|Lactobacillaceae|Pasteurellaceae|Lachnospiraceae|Ruminococcaceae|Tannerellaceae|Bacteroidaceae|Streptococcaceae", feature))

#generate correct q-value
da_dist322_families$qval<-p.adjust(da_dist322_families$pval, method = 'BH')

#write out the results of interest
write_csv(da_dist322_families, "da_results/dist322_family/dist322_targeted_results.csv")

```

## Genus-level
```{r, results='hide', warning=FALSE}
#run full maasiln2 with all comparisons
dist322_genus <- Maaslin2(
  input_data = "genus_tax_features_covid.tsv",
  input_metadata = "covid_metadata_postpand_sample.tsv",
  output = "da_results/dist322_genus",
  min_abundance = 0,
  min_prevalence = 0.5, 
  max_significance = 0.25, 
  analysis_method = "lm", #default
  normalization = "TSS", #default
  transform = "LOG", #default
  fixed_effects = c("dist322"),
  standardize = TRUE
)
```

